---
title: "Clean_AmNat_Edit"
author: "Drew"
date: "2026-09-18"
output: html_document
---
###### ORGANIZATION ######

### Upload packages and data for use
```{r}
### Load Packages
library(MASS)
library(EBSeq)
library(DESeq2)
library(nlme)
library(lme4)
library(lmerTest)
library(piecewiseSEM)
library(rrcov)
library(viridis)
library(pheatmap)

###Pull in Read Count File
ac <- read.csv("Inputs/JT_full_counts.csv")

### Remove Summary Rows leaving only locus read counts
ac<-ac[4:25907,]

### Name rows with Loci identifiers
rownames(ac)<-ac$X
ac<-ac[,2:43]

### Convert to matrix for downstream analysis
ac<-as.matrix(ac)

### Pull in Hormone Measures from previous work (Anderson et al. 2025)
df<-read.csv("Inputs/hormone_data.csv")
df<-df[order(df$ID.1),]

### Get target list ara, arb, er1, er2a, er2b, GER1, PrR, hsd17b4, hsd11b2, srd5a1,srd5a2a,srd5a2b,srd5a3, Aromatase (cyp19a1a), Brain aromatase (cyp19a1b)--Order is the same from this list to gene_id_nb list (e.g. LOC102795905 is ara)
gene_id_nb<-c("LOC102795905","LOC102776319","esr1","esr2a","LOC102788126","gper1","pgr","hsd17b4","hsd11b2","srd5a1","LOC102787127","srd5a2b","srd5a3","LOC102783408","LOC102793447")
```

### Target each brain region's reads 
### Remove genes with low read counts
### Look for outliers
### Normalize read counts for use in Path Analysis (Prediction 4)
```{r}
### Extract POA samples
jt_poa<-ac[,15:28]

### Drop genes with samples with more than half having less than 10 reads
dl<-c()
for(i in 1:length(rownames(jt_poa)))
{
    genei<-jt_poa[i,]
    if((sum(genei>10)<7)=="TRUE")
    {
        dl<-c(dl,i)   
    }
}
jt_poa<-jt_poa[-dl,]

### Normalize read counts for Outlier and PATH Analysis
jt_poa_norm<-vst(jt_poa)

### Get coefficient of variation for each gene
poa_coefv<-c()
for(i in 1:length(rownames(jt_poa_norm)))
{
    genei<-jt_poa_norm[i,]
    cvi<-100*(sd(genei)/mean(genei))
    poa_coefv<-c(poa_coefv,cvi)   
}

### Select groupings based on most variation 
jt_poa_normtopk<-jt_poa_norm[order(-poa_coefv)[1:1000],]
jt_poa_normtop1<-jt_poa_norm[order(-poa_coefv)[1:100],]
jt_poa_normtop5<-jt_poa_norm[order(-poa_coefv)[1:500],]
```

```{r}
### Look for outliers using all genes and genes with highest variation
pg_poa<-PcaGrid(t(jt_poa_norm),kmax = 14, scale=TRUE)
plot(pg_poa)
pg_poak<-PcaGrid(t(jt_poa_normtopk),kmax = 14, scale=TRUE)
plot(pg_poak)
pg_poa5<-PcaGrid(t(jt_poa_normtop5),kmax = 14, scale=TRUE)
plot(pg_poa5)
pg_poa1<-PcaGrid(t(jt_poa_normtop1),kmax = 14, scale=TRUE)
plot(pg_poa1)

### Small Female 2 is the consistent Outlier
```


```{r}
### Extract POA samples dropping outlier (Female Small 2)
jt_poa<-ac[,c(15:18,20:28)]

### Drop genes with samples with more than half having less than 10 reads
dl<-c()
for(i in 1:length(rownames(jt_poa)))
{
    genei<-jt_poa[i,]
    if((sum(genei>10)<7)=="TRUE")
    {
        dl<-c(dl,i)   
    }
}
jt_poa<-jt_poa[-dl,]

### Normalize read counts for PATH Analysis
jt_poa_norm<-vst(jt_poa)

```

### Repeat Process for Vd 
```{r}
### Extract Vd samples
jt_vd<-ac[,29:42]

### Drop genes with samples with more than half having less than 10 reads
dl<-c()
for(i in 1:length(rownames(jt_vd)))
{
    genei<-jt_vd[i,]
    if((sum(genei>10)<7)=="TRUE")
    {
        dl<-c(dl,i)   
    }
}
jt_vd<-jt_vd[-dl,]

### Normalize read counts for Outlier/PATH Analysis
jt_vd_norm<-vst(jt_vd)

### Get coefficient of variation for each gene
vd_coefv<-c()
for(i in 1:length(rownames(jt_vd_norm)))
{
    genei<-jt_vd_norm[i,]
    cvi<-100*(sd(genei)/mean(genei))
    vd_coefv<-c(vd_coefv,cvi)   
}

### Select groupings based on most variation 
jt_vd_normtopk<-jt_vd_norm[order(-vd_coefv)[1:1000],]
jt_vd_normtop1<-jt_vd_norm[order(-vd_coefv)[1:100],]
jt_vd_normtop5<-jt_vd_norm[order(-vd_coefv)[1:500],]
```

```{r}
### Look for outliers using all genes and genes with highest variation
pg_vd<-PcaGrid(t(jt_vd_norm),kmax = 14, scale=TRUE)
plot(pg_vd)
pg_vdk<-PcaGrid(t(jt_vd_normtopk),kmax = 14, scale=TRUE)
plot(pg_vdk)
pg_vd5<-PcaGrid(t(jt_vd_normtop5),kmax = 14, scale=TRUE)
plot(pg_vd5)
pg_vd1<-PcaGrid(t(jt_vd_normtop1),kmax = 14, scale=TRUE)
plot(pg_vd1)
### Larger Female 3 is the consistent Outlier
```

```{r}
### Extract Vd samples dropping outlier (Female Large 3)
jt_vd<-ac[,c(29,30,32:42)]

### Drop genes with samples with more than half having less than 10 reads
dl<-c()
for(i in 1:length(rownames(jt_vd)))
{
    genei<-jt_vd[i,]
    if((sum(genei>10)<7)=="TRUE")
    {
        dl<-c(dl,i)   
    }
}
jt_vd<-jt_vd[-dl,]
### Normalize read counts for PATH Analysis
jt_vd_norm<-vst(jt_vd)

```

### Repeat Process for OT 
```{r}
### Extract OT samples
jt_ot<-ac[,1:14]

### Drop genes with samples with more than half having less than 10 reads
dl<-c()
for(i in 1:length(rownames(jt_ot)))
{
    genei<-jt_ot[i,]
    if((sum(genei>10)<7)=="TRUE")
    {
        dl<-c(dl,i)   
    }
}
jt_ot<-jt_ot[-dl,]

### Normalize read counts for Outlier/PATH Analysis
jt_ot_norm<-vst(jt_ot)

### Get coefficient of variation for each gene
ot_coefv<-c()
for(i in 1:length(rownames(jt_ot_norm)))
{
    genei<-jt_ot_norm[i,]
    cvi<-100*(sd(genei)/mean(genei))
    ot_coefv<-c(ot_coefv,cvi)   
}

### Select groupings based on most variation
jt_ot_normtopk<-jt_ot_norm[order(-ot_coefv)[1:1000],]
jt_ot_normtop1<-jt_ot_norm[order(-ot_coefv)[1:100],]
jt_ot_normtop5<-jt_ot_norm[order(-ot_coefv)[1:500],]
```

```{r}
### Look for outliers using all genes and genes with highest variation
pg_ot<-PcaGrid(t(jt_ot_norm),kmax = 14, scale=TRUE)
plot(pg_ot)
pg_otk<-PcaGrid(t(jt_ot_normtopk),kmax = 14, scale=TRUE)
plot(pg_otk)
pg_ot5<-PcaGrid(t(jt_ot_normtop5),kmax = 14, scale=TRUE)
plot(pg_ot5)
pg_ot1<-PcaGrid(t(jt_ot_normtop1),kmax = 14, scale=TRUE)
plot(pg_ot1)

### Male Big 4 show possible outlier (especially in top 500), but is close to or below cutoff in most conditions. Will keep it
```

### Report Number of Reads kept for analysis
```{r}
### Get count of kept loci and report
poa_count<-length(jt_poa[,1])
vd_count<-length(jt_vd[,1])
ot_count<-length(jt_ot[,1])
countdf<-data.frame(poa_count,vd_count,ot_count)
countdf
```

###### PREDICTION 1 #######

### Model selection for Bayesian Analysis for Whole Gene Set. 
### Note: Dropping the outlier. 
### Gather outputs and generate Table 1
```{r}
### Define Condition Pattern for all samples
JT_Cond<-c("FB","FB","FB","FS","FS","FS","FS","MB","MB","MB","MB","MS","MS","MS")
pat_jtr<-GetPatterns(JT_Cond)

### Normalization for EBSeq--Note this data set is the raw read counts
norm_jt_poa=MedianNorm(jt_poa)

### Test of all fifteen models--Prediction 1
out_jt_poa=EBMultiTest(jt_poa,NgVector=NULL,Conditions=JT_Cond[c(1:4,6:14)], AllParti= pat_jtr, sizeFactors=norm_jt_poa, maxround=100)

### Get summary of models
sumout_jt_poa=GetMultiPP(out_jt_poa)

### Get list table of how many genes fall under each model
tab_raw_poa<-table(sumout_jt_poa$MAP)

### Convert table into percentage of each grouping excluding non-significant genes
tab_per_poa<-table(sumout_jt_poa$MAP)[2:15]*100/(sum(table(sumout_jt_poa$MAP))-table(sumout_jt_poa$MAP)[1])
tab_per_poa<-tab_per_poa[complete.cases(tab_per_poa)]

### Vd Run
###Normalization for EBSeq
norm_jt_vd=MedianNorm(jt_vd)

###Test of all fifteen models
out_jt_vd=EBMultiTest(jt_vd,NgVector=NULL,Conditions=JT_Cond[c(1:2,4:14)], AllParti= pat_jtr, sizeFactors=norm_jt_vd, maxround=100)

###Get summary of models
sumout_jt_vd=GetMultiPP(out_jt_vd)

###Get list table of how many genes fall under each model
tab_raw_vd<-table(sumout_jt_vd$MAP)

### Convert table into percentage of each grouping excluding non-significant genes
tab_per_vd<-table(sumout_jt_vd$MAP)[2:15]*100/(sum(table(sumout_jt_vd$MAP))-table(sumout_jt_vd$MAP)[1])
tab_per_vd<-tab_per_vd[complete.cases(tab_per_vd)]

###OT Run
###Normalization for EBSeq
norm_jt_ot=MedianNorm(jt_ot)

###Test of all fifteen models
out_jt_ot=EBMultiTest(jt_ot,NgVector=NULL,Conditions=JT_Cond, AllParti= pat_jtr, sizeFactors=norm_jt_ot, maxround=100)

###Get summary of models
sumout_jt_ot=GetMultiPP(out_jt_ot)

###Get list table of how many genes fall under each model
tab_raw_ot<-table(sumout_jt_ot$MAP)

### Convert table into percentage of each grouping excluding non-significant genes
tab_per_ot<-table(sumout_jt_ot$MAP)[2:15]*100/(sum(table(sumout_jt_ot$MAP))-table(sumout_jt_ot$MAP)[1])
tab_per_ot<-tab_per_ot[complete.cases(tab_per_ot)]


### Create Dataframe of patterns in order for presentation. Include colname for pattern for merge command
pattern.df<-data.frame(sumout_jt_ot$Patterns)
pattern.df<-data.frame(row.names(pattern.df),pattern.df[,1],pattern.df[,3],pattern.df[,2],pattern.df[,4])
colnames(pattern.df)<-c("Pattern","LF","LM","SF","SM")
pattern.df$Pattern<-c("pattern1","pattern2","pattern3","pattern4","pattern5","pattern6","pattern7","pattern8","pattern9","pattern10","pattern11","pattern12","pattern13","pattern14","pattern15")

### Generate Dataframe of raw counts
tab_raw_all<-merge(data.frame(t(tab_raw_poa))[,2:3],data.frame(t(tab_raw_vd))[,2:3],by="Var2",all=TRUE)
tab_raw_all<-merge(tab_raw_all,data.frame(t(tab_raw_ot))[,2:3],by="Var2",all=TRUE)
colnames(tab_raw_all)<-c("Pattern","POA_Raw","Vd_Raw","OT_Raw")

### Generate Dataframe of percent counts
tab_per_all<-merge(data.frame(t(tab_per_poa))[,2:3],data.frame(t(tab_per_vd))[,2:3],by="Var2",all=TRUE)
tab_per_all<-merge(tab_per_all,data.frame(t(tab_per_ot))[,2:3],by="Var2",all=TRUE)
colnames(tab_per_all)<-c("Pattern","POA_Per","Vd_Per","OT_Per")

### Merge all together
tab_raw<-merge(pattern.df,tab_raw_all,by="Pattern",all=TRUE)
tab_raw_per<-merge(tab_raw,tab_per_all,by="Pattern",all=TRUE)
tab_raw_per<-tab_raw_per[order(match(tab_raw_per$Pattern,c("pattern6","pattern5","pattern3","pattern8","pattern2","pattern4","pattern9","pattern14","pattern13","pattern10","pattern15","pattern7","pattern11","pattern12","pattern1"))),]
tab_raw_per

### Outputs
write.csv(tab_raw_per,"Outputs/Data_For_Table1.csv")
```

### Posterior Probability Graphing 
```{r}
### Grab PP values for genes that were binned in each of the six relevant groupings for each brain region and put them together
### POA
p8_poa<-(sumout_jt_poa$PP[which(rownames(sumout_jt_poa$PP) %in% names(sumout_jt_poa$MAP[which(sumout_jt_poa$MAP == "pattern8")])),c(6,4,8,3,5,2)])
p6_poa<-(sumout_jt_poa$PP[which(rownames(sumout_jt_poa$PP) %in% names(sumout_jt_poa$MAP[which(sumout_jt_poa$MAP == "pattern6")])),c(6,4,8,3,5,2)])
p5_poa<-(sumout_jt_poa$PP[which(rownames(sumout_jt_poa$PP) %in% names(sumout_jt_poa$MAP[which(sumout_jt_poa$MAP == "pattern5")])),c(6,4,8,3,5,2)])
p4_poa<-(sumout_jt_poa$PP[which(rownames(sumout_jt_poa$PP) %in% names(sumout_jt_poa$MAP[which(sumout_jt_poa$MAP == "pattern4")])),c(6,4,8,3,5,2)])
p3_poa<-(sumout_jt_poa$PP[which(rownames(sumout_jt_poa$PP) %in% names(sumout_jt_poa$MAP[which(sumout_jt_poa$MAP == "pattern3")])),c(6,4,8,3,5,2)])
p2_poa<-(sumout_jt_poa$PP[which(rownames(sumout_jt_poa$PP) %in% names(sumout_jt_poa$MAP[which(sumout_jt_poa$MAP == "pattern2")])),c(6,4,8,3,5,2)])
poa_postmap<-rbind(p6_poa,p4_poa,p8_poa,p3_poa,p5_poa,p2_poa)

### Vd
p8_vd<-(sumout_jt_vd$PP[which(rownames(sumout_jt_vd$PP) %in% names(sumout_jt_vd$MAP[which(sumout_jt_vd$MAP == "pattern8")])),c(6,4,8,3,5,2)])
p6_vd<-(sumout_jt_vd$PP[which(rownames(sumout_jt_vd$PP) %in% names(sumout_jt_vd$MAP[which(sumout_jt_vd$MAP == "pattern6")])),c(6,4,8,3,5,2)])
p5_vd<-(sumout_jt_vd$PP[which(rownames(sumout_jt_vd$PP) %in% names(sumout_jt_vd$MAP[which(sumout_jt_vd$MAP == "pattern5")])),c(6,4,8,3,5,2)])
p4_vd<-(sumout_jt_vd$PP[which(rownames(sumout_jt_vd$PP) %in% names(sumout_jt_vd$MAP[which(sumout_jt_vd$MAP == "pattern4")])),c(6,4,8,3,5,2)])
p3_vd<-(sumout_jt_vd$PP[which(rownames(sumout_jt_vd$PP) %in% names(sumout_jt_vd$MAP[which(sumout_jt_vd$MAP == "pattern3")])),c(6,4,8,3,5,2)])
p2_vd<-(sumout_jt_vd$PP[which(rownames(sumout_jt_vd$PP) %in% names(sumout_jt_vd$MAP[which(sumout_jt_vd$MAP == "pattern2")])),c(6,4,8,3,5,2)])
vd_postmap<-rbind(p6_vd,p4_vd,p8_vd,p3_vd,p5_vd,p2_vd)

### OT
p8_ot<-(sumout_jt_ot$PP[which(rownames(sumout_jt_ot$PP) %in% names(sumout_jt_ot$MAP[which(sumout_jt_ot$MAP == "pattern8")])),c(6,4,8,3,5,2)])
p6_ot<-(sumout_jt_ot$PP[which(rownames(sumout_jt_ot$PP) %in% names(sumout_jt_ot$MAP[which(sumout_jt_ot$MAP == "pattern6")])),c(6,4,8,3,5,2)])
p5_ot<-(sumout_jt_ot$PP[which(rownames(sumout_jt_ot$PP) %in% names(sumout_jt_ot$MAP[which(sumout_jt_ot$MAP == "pattern5")])),c(6,4,8,3,5,2)])
p4_ot<-(sumout_jt_ot$PP[which(rownames(sumout_jt_ot$PP) %in% names(sumout_jt_ot$MAP[which(sumout_jt_ot$MAP == "pattern4")])),c(6,4,8,3,5,2)])
p3_ot<-(sumout_jt_ot$PP[which(rownames(sumout_jt_ot$PP) %in% names(sumout_jt_ot$MAP[which(sumout_jt_ot$MAP == "pattern3")])),c(6,4,8,3,5,2)])
p2_ot<-(sumout_jt_ot$PP[which(rownames(sumout_jt_ot$PP) %in% names(sumout_jt_ot$MAP[which(sumout_jt_ot$MAP == "pattern2")])),c(6,4,8,3,5,2)])
ot_postmap<-rbind(p6_ot,p4_ot,p8_ot,p3_ot,p5_ot,p2_ot)

### Heatmap plots for each--POA is for Figure 2
pdf("Outputs/Figure2.pdf")
pheatmap(as.matrix(poa_postmap),color=turbo(100),cluster_cols=FALSE,dendrogram="none",show_rownames = FALSE,treeheight_row=0,labels_col = c("Size","Sex","Fem L","Male L", "Fem S","Male S"), legend_breaks = c(0, .20, .40, .6, .8, .99), main = "POA", cex=1.25,legend_labels = c("0", "0.2", "0.4", "0.6","0.8", "Posterior\nProbability"),border_color=NA)
dev.off()

### Heatmap for Vd for Supplemental 2
pdf("Outputs/Supplemental3.pdf")
pheatmap(as.matrix(vd_postmap),color=turbo(100),cluster_cols=FALSE,dendrogram="none",show_rownames = FALSE,treeheight_row=0,labels_col = c("Size","Sex","Fem L","Male L", "Fem S","Male S"), legend_breaks = c(0, .20, .40, .6, .8, .99), main = "Vd", cex=1.25,legend_labels = c("0", "0.2", "0.4", "0.6","0.8", "Posterior\nProbability"),border_color=NA)
dev.off()

### Heatmap for OT for Supplemental 3
pdf("Outputs/Supplemental4.pdf")
pheatmap(as.matrix(ot_postmap),color=turbo(100),cluster_cols=FALSE,dendrogram="none",show_rownames = FALSE,treeheight_row=0,labels_col = c("Size","Sex","Fem L","Male L", "Fem S","Male S"), legend_breaks = c(0, .20, .40, .6, .8, .99), main = "OT", cex=1.25,legend_labels = c("0", "0.2", "0.4", "0.6","0.8", "Posterior\nProbability"),border_color=NA)
dev.off()
```

### GO Analysis of groups
### Get list of gene identifiers and output for use on the online version of ShinyGO0.86
```{r}
###Get list of genes for GO Analysis
p2.poa<-data.frame(2,names(which(sumout_jt_poa$MAP == "pattern2")))
colnames(p2.poa)<-c("Pattern","Gene")

p3.poa<-data.frame(3,names(which(sumout_jt_poa$MAP == "pattern3")))
colnames(p3.poa)<-c("Pattern","Gene")

p4.poa<-data.frame(4,names(which(sumout_jt_poa$MAP == "pattern4")))
colnames(p4.poa)<-c("Pattern","Gene")

p5.poa<-data.frame(5,names(which(sumout_jt_poa$MAP == "pattern5")))
colnames(p5.poa)<-c("Pattern","Gene")

p6.poa<-data.frame(6,names(which(sumout_jt_poa$MAP == "pattern6")))
colnames(p6.poa)<-c("Pattern","Gene")

p8.poa<-data.frame(8,names(which(sumout_jt_poa$MAP == "pattern8")))
colnames(p8.poa)<-c("Pattern","Gene")

genes.poa<-rbind(p2.poa,p3.poa,p4.poa,p5.poa,p6.poa,p8.poa)

write.csv(genes.poa,"Outputs/Genes_POA.csv")

p2.vd<-data.frame(2,names(which(sumout_jt_vd$MAP == "pattern2")))
colnames(p2.vd)<-c("Pattern","Gene")

p3.vd<-data.frame(3,names(which(sumout_jt_vd$MAP == "pattern3")))
colnames(p3.vd)<-c("Pattern","Gene")

p4.vd<-data.frame(4,names(which(sumout_jt_vd$MAP == "pattern4")))
colnames(p4.vd)<-c("Pattern","Gene")

p5.vd<-data.frame(5,names(which(sumout_jt_vd$MAP == "pattern5")))
colnames(p5.vd)<-c("Pattern","Gene")

p6.vd<-data.frame(6,names(which(sumout_jt_vd$MAP == "pattern6")))
colnames(p6.vd)<-c("Pattern","Gene")

p8.vd<-data.frame(8,names(which(sumout_jt_vd$MAP == "pattern8")))
colnames(p8.vd)<-c("Pattern","Gene")

genes.vd<-rbind(p2.vd,p3.vd,p4.vd,p5.vd,p6.vd,p8.vd)

write.csv(genes.vd,"Outputs/Genes_Vd.csv")

p2.ot<-data.frame(2,names(which(sumout_jt_ot$MAP == "pattern2")))
colnames(p2.ot)<-c("Pattern","Gene")

p3.ot<-data.frame(3,names(which(sumout_jt_ot$MAP == "pattern3")))
colnames(p3.ot)<-c("Pattern","Gene")

p4.ot<-data.frame(4,names(which(sumout_jt_ot$MAP == "pattern4")))
colnames(p4.ot)<-c("Pattern","Gene")

p5.ot<-data.frame(5,names(which(sumout_jt_ot$MAP == "pattern5")))
colnames(p5.ot)<-c("Pattern","Gene")

p6.ot<-data.frame(6,names(which(sumout_jt_ot$MAP == "pattern6")))
colnames(p6.ot)<-c("Pattern","Gene")

p8.ot<-data.frame(8,names(which(sumout_jt_ot$MAP == "pattern8")))
colnames(p8.ot)<-c("Pattern","Gene")

genes.ot<-rbind(p2.ot,p3.ot,p4.ot,p5.ot,p6.ot,p8.ot)

write.csv(genes.ot,"Outputs/Genes_OT.csv")
```

##### PREDICTION 2 and 3 ######

```{r}
###Sample on just our target genes--POA
tar_gen_poa<-jt_poa[which(rownames(jt_poa) %in% gene_id_nb),]

###Model work again on target genes --Prediction 2 and 3
out_jt_tar_poa=EBMultiTest(tar_gen_poa,NgVector=NULL,Conditions=JT_Cond[c(1:4,6:14)], AllParti= pat_jtr, sizeFactors=norm_jt_poa, maxround=100)
sumout_jt_tar_poa=GetMultiPP(out_jt_tar_poa)

### List likely models for genes
print("POA")
sumout_jt_tar_poa$MAP

### Report Posterior Probilities of grouping 8
sumout_jt_tar_poa$PP[,8]

###Sample on just our target genes--Vd
tar_gen_vd<-jt_vd[which(rownames(jt_vd) %in% gene_id_nb),]

###Model work again on target genes--Prediction 2 and 3
out_jt_tar_vd=EBMultiTest(tar_gen_vd,NgVector=NULL,Conditions=JT_Cond[c(1:2,4:14)], AllParti= pat_jtr, sizeFactors=norm_jt_vd, maxround=100)
sumout_jt_tar_vd=GetMultiPP(out_jt_tar_vd)

### List likely models for genes
print("Vd")
sumout_jt_tar_vd$MAP

### Report Posterior Probilities of grouping 4
sumout_jt_tar_vd$PP[,4]

###Sample on just our target genes--OT
tar_gen_ot<-jt_ot[which(rownames(jt_ot) %in% gene_id_nb),]

###Model work again on target genes--Prediction 2 and 3
out_jt_tar_ot=EBMultiTest(tar_gen_ot,NgVector=NULL,Conditions=JT_Cond, AllParti= pat_jtr, sizeFactors=norm_jt_ot, maxround=100)
sumout_jt_tar_ot=GetMultiPP(out_jt_tar_ot)

### List likely models for genes
print("OT")
sumout_jt_tar_ot$MAP
```

### Generate heatmap for Figure 3 and list of z scores for overlay in .ppt
```{r}
###Now we do want to use our target genes. The clunky part above works great here
tar_pp_poa<-sumout_jt_tar_poa$PP[order(rownames(sumout_jt_tar_poa$PP)),c(6,4,8,3,5,2)]
tar_pp_ot<-sumout_jt_tar_ot$PP[order(rownames(sumout_jt_tar_ot$PP)),c(6,4,8,3,5,2)]
tar_pp_vd<-sumout_jt_tar_vd$PP[order(rownames(sumout_jt_tar_vd$PP)),c(6,4,8,3,5,2)]
tar_pp<-merge(tar_pp_poa,tar_pp_vd,by="row.names",all=TRUE)
row.names(tar_pp)<-tar_pp$Row.names
tar_pp<-tar_pp[,2:13]
tar_pp<-merge(tar_pp,tar_pp_ot,by="row.names",all=TRUE)
row.names(tar_pp)<-tar_pp$Row.names
tar_pp<-tar_pp[,2:19]
row.names(tar_pp)<-c("esr1","esr2a","hsd11b2","hsd17b4","ar2","srda5a2a","esr2b","cyp19a1b","ar1","srd5a1","srda3")
tar_pp<-tar_pp[c(4,3,10,6,11,9,5,8,1,2,7),]

pdf("Outputs/Figure3.pdf")
pheatmap(as.matrix(tar_pp),color=turbo(100),cluster_cols=FALSE,cluster_rows=FALSE,dendrogram="none",show_rownames = TRUE,gaps_col=c(6,12),na_col="grey35",treeheight_row=0,labels_col = c("Size","Sex","Fem L","Male L", "Fem S","Male S","Size","Sex","Fem L","Male L", "Fem S","Male S","Size","Sex","Fem L","Male L", "Fem S","Male S"), legend_breaks = c(0, .20, .40, .6, .8, .99), main = "POA                        Vd                           OT",legend_labels = c("0", "0.2", "0.4", "0.6","0.8", "Posterior\nProbability"),cex=1.25,fontsize_row=7,border_color=NA)


xp_poa_tar<-jt_poa_norm[which(rownames(jt_poa_norm) %in% names(sumout_jt_tar_poa$MAP)),]
full<-apply(xp_poa_tar[,1:13],1,mean)
fsd<-apply(xp_poa_tar[,1:13],1,sd)
lfemale<-apply(xp_poa_tar[,1:3],1,mean)
zlf<-(lfemale-full)/fsd
big<-apply(xp_poa_tar[,c(1:3,7:10)],1,mean)
zb<-(big-full)/fsd
sfemale<-apply(xp_poa_tar[,4:6],1,mean)
zsf<-(sfemale-full)/fsd
male<-apply(xp_poa_tar[,7:13],1,mean)
zm<-(male-full)/fsd
lmale<-apply(xp_poa_tar[,7:10],1,mean)
zlm<-(lmale-full)/fsd
smale<-apply(xp_poa_tar[,11:13],1,mean)
zsm<-(smale-full)/fsd
poa_tar_z<-data.frame(zb,zm,zlf,zlm,zsf,zsm)

xp_vd_tar<-jt_vd_norm[which(rownames(jt_vd_norm) %in% names(sumout_jt_tar_vd$MAP)),]
full<-apply(xp_vd_tar[,1:13],1,mean)
fsd<-apply(xp_vd_tar[,1:13],1,sd)
lfemale<-apply(xp_vd_tar[,1:2],1,mean)
zlf<-(lfemale-full)/fsd
big<-apply(xp_vd_tar[,c(1:2,7:10)],1,mean)
zb<-(big-full)/fsd
sfemale<-apply(xp_vd_tar[,3:6],1,mean)
zsf<-(sfemale-full)/fsd
male<-apply(xp_vd_tar[,7:13],1,mean)
zm<-(male-full)/fsd
lmale<-apply(xp_vd_tar[,7:10],1,mean)
zlm<-(lmale-full)/fsd
smale<-apply(xp_vd_tar[,11:13],1,mean)
zsm<-(smale-full)/fsd
vd_tar_z<-data.frame(zb,zm,zlf,zlm,zsf,zsm)

xp_ot_tar<-jt_ot_norm[which(rownames(jt_ot_norm) %in% names(sumout_jt_tar_ot$MAP)),]
full<-apply(xp_ot_tar[,1:14],1,mean)
fsd<-apply(xp_ot_tar[,1:14],1,sd)
lfemale<-apply(xp_ot_tar[,1:3],1,mean)
zlf<-(lfemale-full)/fsd
big<-apply(xp_ot_tar[,c(1:3,8:11)],1,mean)
zb<-(big-full)/fsd
sfemale<-apply(xp_ot_tar[,4:7],1,mean)
zsf<-(sfemale-full)/fsd
male<-apply(xp_ot_tar[,8:14],1,mean)
zm<-(male-full)/fsd
lmale<-apply(xp_ot_tar[,8:11],1,mean)
zlm<-(lmale-full)/fsd
smale<-apply(xp_ot_tar[,12:14],1,mean)
zsm<-(smale-full)/fsd
ot_tar_z<-data.frame(zb,zm,zlf,zlm,zsf,zsm)

zscore<-merge(poa_tar_z,vd_tar_z,by='row.names',all=TRUE)
row.names(zscore)<-zscore$Row.names
zscore<-zscore[,2:13]
zscore<-merge(zscore,ot_tar_z,by='row.names',all=TRUE)
row.names(zscore)<-zscore$Row.names
zscore<-zscore[,2:19]
row.names(zscore)<-c("esr1","esr2a","hsd11b2","hsd17b4","ar2","srda5a2a","esr2b","cyp19a1b","ar1","srd5a1","srda3")
zscore<-zscore[c(4,3,10,6,11,9,5,8,1,2,7),]
colnames(zscore)<-c("POA_Size","POA_Sex","POA_FL","POA_ML","POA_FS","POA_MS","Vd_Size","Vd_Sex","Vd_FL","Vd_ML","Vd_FS","Vd_MS","OT_Size","OT_Sex","OT_FL","OT_ML","OT_FS","OT_MS")
zscore
write.csv(zscore,"Outputs/zscoresforFig3.csv")
```

###### PREDICTION 4 ######

### Imput Hormone and behavior data for use in PATH analysis and Figure 1
```{r}
##Grab normalized read counts from DESeq2 from earlier 
### POA
poa_norm_select<-t(jt_poa_norm[which(rownames(jt_poa_norm) %in% gene_id_nb),])
colnames(poa_norm_select)<-c("hsd11b2","ara","esr2b","cyp19a1b","hsd17b4","esr1","esr2a","arb")
poa.df<-data.frame(df[c(1:4,6:14),c(1:4,9,14:16,19:21,24:27)],poa_norm_select)
poa.df$SexSize<-paste0(poa.df$Sex," ",poa.df$Relative.Size)

### Vd
vd_norm_select<-t(jt_vd_norm[which(rownames(jt_vd_norm) %in% gene_id_nb),])
colnames(vd_norm_select)<-c("hsd11b2","ara","esr2b","cyp19a1b","hsd17b4","esr1","esr2a","arb","srd5a1")
vd.df<-data.frame(df[c(1,2,4:14),c(1:4,9,14:16,19:21,24:27)],vd_norm_select)
vd.df$SexSize<-paste0(vd.df$Sex," ",vd.df$Relative.Size)

### OT
ot_norm_select<-t(jt_ot_norm[which(rownames(jt_ot_norm) %in% gene_id_nb),])
colnames(ot_norm_select)<-c("hsd11b2","hsd17b4","srd5a3","srd5a2a","srd5a1")
ot.df<-data.frame(df[,c(1:4,9,14:16,19:21,24:27)],ot_norm_select)
ot.df$SexSize<-paste0(ot.df$Sex," ",ot.df$Relative.Size)
ot.df$SexSize<-factor(ot.df$SexSize,levels=c("Male Bigger","Male Smaller","Female Bigger","Female Smaller"))

```

###Use OT values for hormones and behaviors for Figure 1 since there are no dropped outliers.
###Figure 1 (Image added in .ppt)
```{r}
pdf("Outputs/Figure1.pdf")

par(mfrow=c(2,3))

par(mar=c(.5,5,.75,1)) 
boxplot(Out~SexSize,ot.df,col=c("#86BBB8","#86BBB8","#F6AE2D","#F6AE2D"),xaxt="n",cex.axis=1.25,cex.lab=1.75, xlab="",ylab="Time Out of Nest (s)",ylim=c(0,1250))

par(mar=c(.5,6,.75,0))
boxplot(Out_Dif~SexSize,ot.df,col=c("#86BBB8","#86BBB8","#F6AE2D","#F6AE2D"),xaxt="n",cex.axis=1.25,cex.lab=1.75, xlab="",ylab="Difference in Time Out\nof Nest (s) with Partner")

par(mar=c(.5,5,.75,0.5))
boxplot(X11KT~SexSize,ot.df,col=c("#86BBB8","#86BBB8","#F6AE2D","#F6AE2D"),xaxt="n",cex.axis=1.25,cex.lab=1.75, xlab="",ylab="11-Ketotestosterone (pg/g/hr)")
axis(1,at=c(1:4),labels = c("Larger\nMale","Smaller\nMale","Larger\nFemale","Smaller\nFemale"),cex.axis=0.75,tick=FALSE)

par(mar=c(2.5,5,.75,1))
boxplot(asinh(Shim)~SexSize,ot.df,col=c("#86BBB8","#86BBB8","#F6AE2D","#F6AE2D"),xaxt="n",cex.axis=1.25,cex.lab=1.75, xlab="",ylab="Number of Shimmies")
axis(1,at=c(1:4),labels = c("Larger\nMale","Smaller\nMale","Larger\nFemale","Smaller\nFemale"),cex.axis=0.75,tick=FALSE)

par(mar=c(2.5,6,.75,0))
boxplot(asinh(Shim_Dif)~SexSize,ot.df,col=c("#86BBB8","#86BBB8","#F6AE2D","#F6AE2D"),xaxt="n",cex.axis=1.25,cex.lab=1.75, xlab="",ylab="Difference in Number\nof Shimmies with Partner")
axis(1,at=c(1:4),labels = c("Larger\nMale","Smaller\nMale","Larger\nFemale","Smaller\nFemale"),cex.axis=0.75,tick=FALSE)
dev.off()
```

### piecewise SEM model Approach
### Now with can do a path analysis to account for variation as we move through the hormone synthesis all the way to the receptors
### Note we can now account for all variables, but effects are assumed to be one-directional
### First model assumes upstream features directly affect the next step in hormone path (e.g. T-->aromatase)
### Saved outputs can be combined to generate Supplemental 1
```{r}
### Scale needed data to ensure values are comparable.
poa.df2<-data.frame(poa.df[,1:11],scale(poa.df[,12:23]),poa.df[,24])

### Original model and assumptions
appsem1.poa<-psem(glm(hsd17b4~Pr,family="gaussian",data=poa.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=poa.df2),glm(arb~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df2),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df2),glm(cyp19a1b~T+hsd17b4,family="gaussian",poa.df2),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df2),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df2),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df2),glm(App~ara+arb+esr1+esr2a+esr2b,family="poisson",poa.df2),data=poa.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(appsem1.poa)[5]

### Look for direct relationships not considered in model that should be added
df.apppoa1<-data.frame(summary(appsem1.poa)[3])
df.apppoa1[which(df.apppoa1$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves
appsem2.poa<-psem(glm(hsd17b4~Pr+T,family="gaussian",data=poa.df2),glm(hsd11b2~hsd17b4+T+X11KT,family="gaussian",data=poa.df2),glm(arb~T+hsd11b2+X11KT+hsd17b4+Pr,family="gaussian",data=poa.df2),glm(ara~T+hsd11b2+hsd17b4+X11KT,family="gaussian",data=poa.df2),glm(cyp19a1b~T+hsd17b4+ara,family="gaussian",poa.df2),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df2),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df2),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df2),glm(App~ara+arb+esr1+esr2a+esr2b+hsd11b2+T+Pr,family="poisson",poa.df2),data=poa.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(appsem2.poa)[5]

### Model good fit now report correlation values
df.apppoa2<-data.frame(summary(appsem2.poa)[7])
df.apppoa2

write.csv(df.apppoa2,"Outputs/PATH_POA_app.csv")
```

###piecewise SEM model Out of Nest
```{r}
### normalize out of nest data following Anderson et al., 2025
poa.df2$outs<-asin(poa.df2$Out/1200)

### Original model and assumptions
outsem1.poa<-psem(glm(hsd17b4~Pr,family="gaussian",data=poa.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=poa.df2),glm(arb~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df2),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df2),glm(cyp19a1b~T+hsd17b4,family="gaussian",poa.df2),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df2),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df2),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df2),glm(outs~ara+arb+esr1+esr2a+esr2b,family="gaussian",poa.df2),data=poa.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(outsem1.poa)[5]

### Technically above 0.05, but need see if the upstream factors can be added in to improve model
### Look for direct relationships not considered in model that should be added
df.outpoa1<-data.frame(summary(outsem1.poa)[3])
df.outpoa1[which(df.outpoa1$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves

outsem2.poa<-psem(glm(hsd17b4~Pr+T,family="gaussian",data=poa.df2),glm(hsd11b2~hsd17b4+T+X11KT,family="gaussian",data=poa.df2),glm(arb~T+hsd11b2+X11KT+Pr+hsd17b4,family="gaussian",data=poa.df2),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df2),glm(cyp19a1b~T+hsd17b4+ara,family="gaussian",poa.df2),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df2),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df2),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df2),glm(outs~ara+arb+esr1+esr2a+esr2b,family="gaussian",poa.df2),data=poa.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(outsem2.poa)[5]

### Model good fit now report correlation values
df.outpoa2<-data.frame(summary(outsem2.poa)[7])
df.outpoa2

write.csv(df.outpoa2,"Outputs/PATH_POA_out.csv")
```

###piecewise SEM model Shimmy
```{r}
### Original model and assumptions
shimsem1.poa<-psem(glm(hsd17b4~Pr,family="gaussian",data=poa.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=poa.df2),glm(arb~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df2),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df2),glm(cyp19a1b~T+hsd17b4,family="gaussian",poa.df2),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df2),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df2),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df2),glm(Shim~ara+arb+esr1+esr2a+esr2b,family="poisson",poa.df2),data=poa.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(shimsem1.poa)[5]

### Look for direct relationships not considered in model that should be added
df.shimpoa1<-data.frame(summary(shimsem1.poa)[3])
df.shimpoa1[which(df.shimpoa1$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves

shimsem2.poa<-psem(glm(hsd17b4~Pr+T,family="gaussian",data=poa.df2),glm(hsd11b2~hsd17b4+T+X11KT,family="gaussian",data=poa.df2),glm(arb~T+hsd11b2+X11KT+hsd17b4+Pr,family="gaussian",data=poa.df2),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df2),glm(cyp19a1b~T+hsd17b4+ara,family="gaussian",poa.df2),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df2),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df2),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df2),glm(Shim~ara+arb+esr1+esr2a+esr2b+hsd11b2+E2+X11KT+T,family="poisson",poa.df2),data=poa.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(shimsem2.poa)[5]

### Model good fit now report correlation values
df.shimpoa2<-data.frame(summary(shimsem2.poa)[7])
df.shimpoa2

write.csv(df.shimpoa2,"Outputs/PATH_POA_shim.csv")
```
### Not sure how to visually represent path results in R plotting. Used coefficients x5 for line widths and made figure in Powerpoint

###Vd

###piecewise SEM model Approach
```{r}
### Scale needed data to ensure values are comparable.
vd.df2<-data.frame(vd.df[,1:11],scale(vd.df[,12:24]))

### Original model and assumptions
appsem1.vd<-psem(glm(hsd17b4~Pr,family="gaussian",data=vd.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=vd.df2),glm(srd5a1~hsd17b4+T,family="gaussian",vd.df2),glm(arb~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(ara~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(cyp19a1b~T+hsd17b4,family="gaussian",vd.df2),glm(esr1~E2+cyp19a1b,family="gaussian",vd.df2),glm(esr2a~cyp19a1b+E2,family="gaussian",vd.df2),glm(esr2b~cyp19a1b+E2,family="gaussian",vd.df2),glm(App~ara+arb+esr1+esr2a+esr2b,family="poisson",vd.df2),data=vd.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(appsem1.vd)[5]

### Look for direct relationships not considered in model that should be added
df.appvd1<-data.frame(summary(appsem1.vd)[3])
df.appvd1[which(df.appvd1$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves

appsem2.vd<-psem(glm(hsd17b4~Pr,family="gaussian",data=vd.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=vd.df2),glm(srd5a1~hsd17b4+T,family="gaussian",vd.df2),glm(arb~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(ara~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(cyp19a1b~T+hsd17b4+Pr,family="gaussian",vd.df2),glm(esr1~E2+cyp19a1b+hsd17b4,family="gaussian",vd.df2),glm(esr2a~cyp19a1b+E2+Pr,family="gaussian",vd.df2),glm(esr2b~cyp19a1b+E2+hsd17b4,family="gaussian",vd.df2),glm(App~ara+arb+esr1+esr2a+esr2b+srd5a1+hsd17b4+E2+T+Pr,family="poisson",vd.df2),data=vd.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(appsem2.vd)[5]

### Look for direct relationships not considered in model that should be added
df.appvd2<-data.frame(summary(appsem2.vd)[3])
df.appvd2[which(df.appvd2$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves--Didn't add ara by cyp19a1b since the direction is wrong. Didn't add cyp19a1b or 11KT to App due to to0 many factors

appsem3.vd<-psem(glm(hsd17b4~Pr,family="gaussian",data=vd.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=vd.df2),glm(srd5a1~hsd17b4+T,family="gaussian",vd.df2),glm(arb~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(ara~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(cyp19a1b~T+hsd17b4+Pr,family="gaussian",vd.df2),glm(esr1~E2+cyp19a1b+hsd17b4+Pr,family="gaussian",vd.df2),glm(esr2a~cyp19a1b+E2+Pr,family="gaussian",vd.df2),glm(esr2b~cyp19a1b+E2+hsd17b4+Pr,family="gaussian",vd.df2),glm(App~ara+arb+esr1+esr2a+esr2b+srd5a1+hsd17b4+E2+T+Pr,family="poisson",vd.df2),data=vd.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(appsem3.vd)[5]

### Model good fit now report correlation values
df.appvd3<-data.frame(summary(appsem3.vd)[7])
df.appvd3

write.csv(df.appvd3,"Outputs/PATH_Vd_app.csv")
```


###Shimmy
```{r}
### Original model and assumptions
shimsem1.vd<-psem(glm(hsd17b4~Pr,family="gaussian",data=vd.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=vd.df2),glm(srd5a1~hsd17b4+T,family="gaussian",vd.df2),glm(arb~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(ara~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(cyp19a1b~T+hsd17b4,family="gaussian",vd.df2),glm(esr1~E2+cyp19a1b,family="gaussian",vd.df2),glm(esr2a~cyp19a1b+E2,family="gaussian",vd.df2),glm(esr2b~cyp19a1b+E2,family="gaussian",vd.df2),glm(Shim~ara+arb+esr1+esr2a+esr2b,family="poisson",vd.df2),data=vd.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(shimsem1.vd)[5]

### Look for direct relationships not considered in model that should be added
df.shimvd1<-data.frame(summary(shimsem1.vd)[3])
df.shimvd1[which(df.shimvd1$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves

shimsem2.vd<-psem(glm(hsd17b4~Pr,family="gaussian",data=vd.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=vd.df2),glm(srd5a1~hsd17b4+T,family="gaussian",vd.df2),glm(arb~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(ara~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(cyp19a1b~T+hsd17b4+Pr,family="gaussian",vd.df2),glm(esr1~E2+cyp19a1b+hsd17b4,family="gaussian",vd.df2),glm(esr2a~cyp19a1b+E2+Pr,family="gaussian",vd.df2),glm(esr2b~cyp19a1b+E2+hsd17b4,family="gaussian",vd.df2),glm(Shim~ara+arb+esr1+esr2a+esr2b+srd5a1+hsd11b2+hsd17b4+E2+T,family="poisson",vd.df2),data=vd.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(shimsem2.vd)[5]

### Look for direct relationships not considered in model that should be added
df.shimvd2<-data.frame(summary(shimsem2.vd)[3])
df.shimvd2[which(df.shimvd2$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves--Didn't add ara by cyp19a1b since the direction is wrong.

shimsem3.vd<-psem(glm(hsd17b4~Pr,family="gaussian",data=vd.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=vd.df2),glm(srd5a1~hsd17b4+T,family="gaussian",vd.df2),glm(arb~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(ara~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(cyp19a1b~T+hsd17b4+Pr,family="gaussian",vd.df2),glm(esr1~E2+cyp19a1b+hsd17b4+Pr,family="gaussian",vd.df2),glm(esr2a~cyp19a1b+E2+Pr,family="gaussian",vd.df2),glm(esr2b~cyp19a1b+E2+hsd17b4+Pr,family="gaussian",vd.df2),glm(Shim~ara+arb+esr1+esr2a+esr2b+srd5a1+hsd11b2+hsd17b4+E2+T,family="poisson",vd.df2),data=vd.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(shimsem3.vd)[5]

### Model good fit now report correlation values
df.shimvd3<-data.frame(summary(shimsem3.vd)[7])
df.shimvd3

write.csv(df.shimvd3,"Outputs/PATH_Vd_shim.csv")
```


### Out of Nest
```{r}
### normalize out of nest data following Anderson et al., 2025
vd.df2$outs<-asin(vd.df2$Out/1200)

### Original model and assumptions
outsem1.vd<-psem(glm(hsd17b4~Pr,family="gaussian",data=vd.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=vd.df2),glm(srd5a1~hsd17b4+T,family="gaussian",vd.df2),glm(arb~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(ara~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(cyp19a1b~T+hsd17b4,family="gaussian",vd.df2),glm(esr1~E2+cyp19a1b,family="gaussian",vd.df2),glm(esr2a~cyp19a1b+E2,family="gaussian",vd.df2),glm(esr2b~cyp19a1b+E2,family="gaussian",vd.df2),glm(outs~ara+arb+esr1+esr2a+esr2b,family="gaussian",vd.df2),data=vd.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(outsem1.vd)[5]

### Look for direct relationships not considered in model that should be added
df.outvd1<-data.frame(summary(outsem1.vd)[3])
df.outvd1[which(df.outvd1$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves

outsem2.vd<-psem(glm(hsd17b4~Pr,family="gaussian",data=vd.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=vd.df2),glm(srd5a1~hsd17b4+T,family="gaussian",vd.df2),glm(arb~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(ara~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(cyp19a1b~T+hsd17b4+Pr,family="gaussian",vd.df2),glm(esr1~E2+cyp19a1b+hsd17b4,family="gaussian",vd.df2),glm(esr2a~cyp19a1b+E2+Pr,family="gaussian",vd.df2),glm(esr2b~cyp19a1b+E2+hsd17b4,family="gaussian",vd.df2),glm(outs~ara+arb+esr1+esr2a+esr2b+cyp19a1b,family="gaussian",vd.df2),data=vd.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(outsem2.vd)[5]

### Look for direct relationships not considered in model that should be added
df.outvd2<-data.frame(summary(outsem2.vd)[3])
df.outvd2[which(df.outvd2$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves--Didn't add ara by cyp19a1b since the direction is wrong.

outsem3.vd<-psem(glm(hsd17b4~Pr,family="gaussian",data=vd.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=vd.df2),glm(srd5a1~hsd17b4+T,family="gaussian",vd.df2),glm(arb~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(ara~T+hsd11b2+X11KT+srd5a1+hsd17b4,family="gaussian",data=vd.df2),glm(cyp19a1b~T+hsd17b4+Pr,family="gaussian",vd.df2),glm(esr1~E2+cyp19a1b+hsd17b4+Pr+srd5a1,family="gaussian",vd.df2),glm(esr2a~cyp19a1b+E2+Pr,family="gaussian",vd.df2),glm(esr2b~cyp19a1b+E2+hsd17b4+Pr,family="gaussian",vd.df2),glm(outs~ara+arb+esr1+esr2a+esr2b+cyp19a1b+hsd17b4,family="gaussian",vd.df2),data=vd.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(outsem3.vd)[5]

### Model good fit now report correlation values
df.outvd3<-data.frame(summary(outsem3.vd)[7])
df.outvd3

write.csv(df.outvd3,"Outputs/PATH_Vd_out.csv")
```


###piecewise SEM model Approach OT
```{r}
### Scale needed data to ensure values are comparable.
ot.df2<-data.frame(ot.df[,1:11],scale(ot.df[,12:20]))

### Original model and assumptions
appsem1.ot<-psem(glm(hsd17b4~Pr,family="gaussian",data=ot.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=ot.df2),glm(srd5a1~T+hsd17b4,family="gaussian",data=ot.df2),glm(srd5a2a~T+hsd17b4,family="gaussian",data=ot.df2),glm(srd5a3~T+hsd17b4,family="gaussian",data=ot.df2),glm(App~1,family="poisson",ot.df2),data=ot.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(appsem1.ot)[5]

### Model good fit now report correlation values
df.appot1<-data.frame(summary(appsem1.ot)[7])
df.appot1

## no significance anywhere
write.csv(df.appot1,"Outputs/PATH_OT_app.csv")
```

###piecewise SEM model Out of Nest
```{r}
### normalize out of nest data following Anderson et al., 2025
ot.df2$outs<-asin(ot.df2$Out/1200)

### Original model and assumptions
outsem1.ot<-psem(glm(hsd17b4~Pr,family="gaussian",data=ot.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=ot.df2),glm(srd5a1~T+hsd17b4,family="gaussian",data=ot.df2),glm(srd5a2a~T+hsd17b4,family="gaussian",data=ot.df2),glm(srd5a3~T+hsd17b4,family="gaussian",data=ot.df2),glm(outs~1,family="gaussian",ot.df2),data=ot.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(outsem1.ot)[5]

### Model good fit now report correlation values
df.outot1<-data.frame(summary(outsem1.ot)[7])
df.outot1

## No significant
write.csv(df.outot1,"Outputs/PATH_OT_out.csv")
```


###piecewise SEM model Shimmy
```{r}
### Original model and assumptions
shimsem1.ot<-psem(glm(hsd17b4~Pr,family="gaussian",data=ot.df2),glm(hsd11b2~hsd17b4+T,family="gaussian",data=ot.df2),glm(srd5a1~T+hsd17b4,family="gaussian",data=ot.df2),glm(srd5a2a~T+hsd17b4,family="gaussian",data=ot.df2),glm(srd5a3~T+hsd17b4,family="gaussian",data=ot.df2),glm(Shim~1,family="poisson",ot.df2),data=ot.df2)

### Test for model fit (Fisher.C p-value > 0.05)
summary(shimsem1.ot)[5]

### Model good fit now report correlation values
df.shimot1<-data.frame(summary(shimsem1.ot)[7])
df.shimot1

write.csv(df.shimot1,"Outputs/PATH_OT_shim.csv")
```

### Concern from Reviews of low sample size and if any group is driving relationship. 
### Chose to look at POA and large females as lots of evidence these females are doing something unique with these genes/hormones. 
### Removed large females to see how that alters PATH results.  
### Output is include in Supplemental 4

```{r}
### Scale needed data to ensure values are comparable.
poa.df3<-data.frame(poa.df[4:13,1:11],scale(poa.df[4:13,12:23]),poa.df[4:13,24])

### Original model and assumptions
appsem1.poaf<-psem(glm(hsd17b4~Pr,family="gaussian",data=poa.df3),glm(hsd11b2~hsd17b4+T,family="gaussian",data=poa.df3),glm(arb~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(cyp19a1b~T+hsd17b4,family="gaussian",poa.df3),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df3),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df3),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df3),glm(App~ara+arb+esr1+esr2a+esr2b,family="poisson",poa.df3),data=poa.df3)

### Test for model fit (Fisher.C p-value > 0.05)
summary(appsem1.poaf)[5]

### Look for direct relationships not considered in model that should be added
df.apppoa1b<-data.frame(summary(appsem1.poaf)[3])
df.apppoa1b[which(df.apppoa1b$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves

appsem2.poaf<-psem(glm(hsd17b4~Pr,family="gaussian",data=poa.df3),glm(hsd11b2~hsd17b4+T,family="gaussian",data=poa.df3),glm(arb~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(cyp19a1b~T+hsd17b4+ara,family="gaussian",poa.df3),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df3),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df3),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df3),glm(App~ara+arb+esr1+esr2a+esr2b+hsd11b2+cyp19a1b,family="poisson",poa.df3),data=poa.df3)

### Test for model fit (Fisher.C p-value > 0.05)
summary(appsem2.poaf)[5]

### Model good fit now report correlation values
df.apppoa2b<-data.frame(summary(appsem2.poaf)[7])
df.apppoa2b

write.csv(df.apppoa2b,"Outputs/PATH_POA_Drop_app.csv")
```

###piecewise SEM model Out of Nest
```{r}
### normalize out of nest data following Anderson et al., 2025
poa.df3$outs<-asin(poa.df3$Out/1200)

### Original model and assumptions
outsem1.poaf<-psem(glm(hsd17b4~Pr,family="gaussian",data=poa.df3),glm(hsd11b2~hsd17b4+T,family="gaussian",data=poa.df3),glm(arb~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(cyp19a1b~T+hsd17b4,family="gaussian",poa.df3),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df3),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df3),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df3),glm(outs~ara+arb+esr1+esr2a+esr2b,family="gaussian",poa.df3),data=poa.df3)

### Test for model fit (Fisher.C p-value > 0.05)
summary(outsem1.poaf)[5]

### Technically above 0.05, but need see if the upstream factors can be added in to improve model
### Look for direct relationships not considered in model that should be added
df.outpoa1b<-data.frame(summary(outsem1.poaf)[3])
df.outpoa1b[which(df.outpoa1b$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves

outsem2.poaf<-psem(glm(hsd17b4~Pr,family="gaussian",data=poa.df3),glm(hsd11b2~hsd17b4+T,family="gaussian",data=poa.df3),glm(arb~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(cyp19a1b~T+hsd17b4+ara,family="gaussian",poa.df3),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df3),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df3),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df3),glm(outs~ara+arb+esr1+esr2a+esr2b,family="gaussian",poa.df3),data=poa.df3)

### Test for model fit (Fisher.C p-value > 0.05)
summary(outsem2.poaf)[5]

### Model good fit now report correlation values
df.outpoa2b<-data.frame(summary(outsem2.poaf)[7])
df.outpoa2b

write.csv(df.outpoa2b,"Outputs/PATH_POA_Drop_out.csv")
```

###piecewise SEM model Shimmy
```{r}
### Original model and assumptions
shimsem1.poaf<-psem(glm(hsd17b4~Pr,family="gaussian",data=poa.df3),glm(hsd11b2~hsd17b4+T,family="gaussian",data=poa.df3),glm(arb~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(cyp19a1b~T+hsd17b4,family="gaussian",poa.df3),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df3),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df3),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df3),glm(Shim~ara+arb+esr1+esr2a+esr2b,family="poisson",poa.df3),data=poa.df3)

### Test for model fit (Fisher.C p-value > 0.05)
summary(shimsem1.poaf)[5]

### Technically above 0.05, but need see if the upstream factors can be added in to improve model
### Look for direct relationships not considered in model that should be added
df.shimpoa1b<-data.frame(summary(shimsem1.poaf)[3])
df.shimpoa1b[which(df.shimpoa1b$dTable.P.Value <= 0.05),]

### Add in those relationships and see if model improves

shimsem2.poaf<-psem(glm(hsd17b4~Pr,family="gaussian",data=poa.df3),glm(hsd11b2~hsd17b4+T,family="gaussian",data=poa.df3),glm(arb~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(ara~T+hsd11b2+X11KT+hsd17b4,family="gaussian",data=poa.df3),glm(cyp19a1b~T+hsd17b4+ara,family="gaussian",poa.df3),glm(esr1~E2+cyp19a1b,family="gaussian",poa.df3),glm(esr2a~cyp19a1b+E2,family="gaussian",poa.df3),glm(esr2b~cyp19a1b+E2,family="gaussian",poa.df3),glm(Shim~ara+arb+esr1+esr2a+esr2b+hsd11b2+cyp19a1b,family="poisson",poa.df3),data=poa.df3)

### Test for model fit (Fisher.C p-value > 0.05)
summary(shimsem2.poaf)[5]

### Model good fit now report correlation values
df.shimpoa2b<-data.frame(summary(shimsem2.poaf)[7])
df.shimpoa2b

write.csv(df.shimpoa2b,"Outputs/PATH_POA_Drop_shim.csv")
```